Many people have been involved in the process of working on this thesis, and I would like to take the opportunity to thank them.

First and foremost, I would like to thank my supervisor Jochen Schieck for encouraging me to pursue this path and for the opportunity to work at CERN. His support and trust have been invaluable throughout this journey.

Having worked on this thesis far from my home university, I am grateful to my CERN supervisor Andreas Salzburger for his calm, thoughtful, and far-sighted way of leading the tracking group I found myself in at CERN. His continuous trust, support, and expertise have been instrumental in the successful completion of this work.

I also want to thank Paul and Noemi, for their co-supervision with constant help, patience, fruitful discussions, and feedback on my work, but also on a personal level as collegues, office mates, and friends. They individually guided me through various aspects of this thesis. Apart from many other things I thank Paul for reviewing hundreds of my code contributions and Noemi for proofreading large parts of this thesis.

At CERN, I had the chance to work with smart, talented, and supportive colleagues. In particular, I want to thank Benjamin for countless in-depth discussions, reviewing a significant amount of my code, and for cheering me up with his down-to-earth life wisdom and his way of spinning things positively. I also want to thank Pierfrancesco and Valentina for their feedback on my work (and slides) and for challenging me with different topics, questions, and discussions that helped me to sharpen my understanding of many aspects of this thesis.

% On a more general level, I want to thank the whole ACTS community and ATLAS collaboration for the welcoming and open atmosphere that made it easy for me to get involved.

I want to thank my family and friends for supporting me under all circumstances, especially given the increased distance after moving to the Geneva area.

Last but not least, I want to thank my partner Samira for always supporting and encouraging me. This includes the long journeys from Vienna to Geneva and back, moving to a different country, and enduring me working countless weekends and late nights on this thesis.
