Lots of developments were done during the course of this thesis. The previous chapters framed the big topics of these developments. They included developments on the propagation, the support for dense material during propagation, the improvements to the track finding, and developments on 4D tracking. A total of 581 pull requests have been merged to Acts \texttt{main} during the course of this thesis, from May 2022 to April 2025. Some of them were small bug fixes, others large scale refactors, and some were new features. All of them contributed to the overall goal of improving Acts by accomodating needs from experiments, making it more experiment independent, adaptable, maintainable, and efficient. Software is not a stationary object due to its changing environment and requires continuous maintainance and improvements.

One of the central features conceptualized and implemented during this thesis was the introduction of the particle hypothesis. It is a crucial ingredient for reconstruction and unavoidable for an experiment independent track reconstruction toolkit.

A first version of a greedy ambiguity resolver was implemented. It can be a crucial part of the track finding and improve the physics performance of downstream algorithms like vertexing. Future work can be based on this implementation and further improve its performance and flexibility.

The navigation was improved and hardened to be more robust. This has direct implications to track finding and fitting, which rely on the navigation to find relevant sensors. In physics terms this can make a difference for the track efficiency, purity, and fake rate.

The stepping performance was improved substantially with the implementation of the \texttt{SympyStepper}. The performance improvements are crucial for track finding and fitting, as stepping is a central part of the track propagation. The approach of using symbolic mathematics to generate code might have further applications in the future.

The track finding has been effectively redesigned and modularized. The new design provides building blocks for track finding solutions and puts the user in control of the finding procedure. This allows for more aggressive optimizations and experiment specific solutions.

The Examples framework has been improved over several iterations. The main purpose for these developments was to enable detailed studies on physics performance. The Examples framework is now more flexible and can be used for a wide range of studies.

Not all ideas have been followed up and implemented. This is summarized in the following sections: shortcoming during the work on this thesis and proposals for future developments.

\section{Shortcomings}

The computational performance of the \texttt{SympyStepper} is still not on par with the \texttt{AtlasStepper}. Stepping plays a vital role in tracking and its performance directly affects the performance of the track finding and fitting. Closing this gap could reduce the CPU time during track finding by a few percent.

The navigation could be generalized further by combining and integrating the target surface approach with the external surfaces. Currently, the target approach is handled by a propagation \texttt{Actor} which behaves like a second navigator. This is necessary because the target surface is usually not part of the geometry. The same is true for external surfaces. The navigation could be further improved by providing a more general approach to handle these cases.

Surface ambiguity arises when multiple surfaces are close to each other. The navigation has to decide which one to target, since both might be compatible with the estimated trajectory and error. This effect can lead to inefficiencies during track finding.

During track finding, edge holes are not yet handled properly and can lead to inefficiencies. Handling multiple compatible surfaces is not implemented which can also result in inefficiencies if relevant. Additionally, Acts could provide a modular implementation of the \texttt{TrackStateCreator} which generalizes to different experiments and allows for detailed customization.

A solution to stepping in dense material has beed proposed and implemented. In comparison to Geant4, not all material effects seem to be taken into account or are not taken into account the same way. This can lead to discrepancies between simulation and reconstruction which can affect the physics performance of the track finding and fitting.

4D tracking has made a big step forward in Acts. But there is still only one vertex seeder implementation that supports time information. To validate its performance, the truth based seeder can be used for now. Adaptation of the analytical gaussian track density seeder to support time information is not yet implemented.

\section{Proposals for future developments}

Experiment independence is a key feature of Acts. To further improve this, the track parameterization should be customizable. This includes the potential unused time parameter. Similarly, the units and constants should be customizable to give experiments more control over the numerical precision. This is a fairly long term goal and potentially requires large scale refactors. But keeping it in mind might open up new possibilities for the future.

The caller and callee inversion for propagation dependent algorithms has the potential to make Acts more readable and maintainable. This will most likely not impact performance but can make the code more accessible to new developers. Additionally, debugging and profiling should be easier with this approach. This can be pioneered with the \texttt{KalmanFitter} and later discussed with the community.

The geometry identification in Acts is ad-hoc and not flexible. As such, experiments have to adapt to the Acts geometry identification or create a mapping from their own scheme. The main use case for this identification is to map the hierarchical structure of the detector which then allows to identify important subdetectors, like pixel or strip detectors. A flexible approach could allow experiments to define their own hierarchy and identification scheme.

The usage of virtual interfaces is not consistent throughout Acts. While the geometry and magnetic field are fully virtual, the propagator and track EDM use template based interfaces instead. The main argument to choose the template based approach was performance. However, the performance impact of virtual interfaces might be negligible with the current workload and the argument falls short looking at the virtual geometry implementation. Virtual interfaces make the code more readable and maintainable. This should be re-evaluated and potentially changed.

Currently, the propagation can only be started from bound parameters and if final parameters are requested, only bound parameters are returned. Using free track parameters for either one, start or end, would be an equally valid choice. The main reason for bound parameters is that algorithms like the \texttt{KalmanFitter} require them to handle bound measurements. Changing this would require a flexible handling of the transport Jacobian and covariance matrix. Free parameters might be preferable during vertexing and for measurements that are not bound to a surface.
