Navigation is a core concept in tracking. It is essential for efficient track propagtion through the detector. The objective of the navigation is to find the next surface the trajectory will potentially intersect with, given the current position, direction, and tracking geometry. The stepper will then transport the track parameters towards the surface using an estimated path length.

The navigation has to be both fast and accurate: fast, because it lives in a computationally hot path for high throughput algorithms like fast track simulation, combinatorial track finding, and track fitting; accurate, because it will ultimately dictate which surfaces will be looked at. Navigation inaccuracies will directly influence the track finding efficiency and the quality of found and fitted tracks. These two goals are usually in conflict and one has to find a sweet spot to achieve good performance for both, physics and compute time.

A good understanding of the navigation is essential to understand the performance of the track finding algorithm in ACTS. Since track finding performance was a main focus of this thesis, and the navigation turned out to be a limiting factor, it has also been subject to several improvements.

This chapter will give an overview of the navigation in ACTS, its current implementation, and the improvements it was subject to. First, general subtleties and edge cases of navigation are discussed, which are not specific to ACTS but are relevant for any algorithm that deals with track propagation. Then, a streamlined navigation model is presented, which includes a robust intersection handling for cylinders, the removal of path and overstep limit, and the decoupling of the navigation from the stepping. After that, special navigator implementations in ACTS are discussed, the \texttt{DirectNavigator} and the try-all navigators. Finally, a conceptualized solution for surface ambiguity is presented.

\section{Navigation subtleties and edge cases}

Naively one would assume that navigation is a simple task and generally a solved problem. This is true for straight tracks where we can accurately predict the next surface the trajectory will intersect with. In such a case one just needs to check the computational performance of the navigation. But since tracking deals with curved trajectories, the navigation becomes a harder problem. Given a homogenous magnetic field and ignoring any interactions, our trajectories become helices. Checking for an intersection between a helix and a surface does not have a direct solution, even for plane surfaces, and comes with additional computational cost. Given inhomogenous magnetic fieds and interactions, which is the general case ACTS has to handle, even an helix would only be an approximation.

The generic navigation implementation in ACTS, \texttt{Navigator}, approximates trajectories locally with straight lines and uses them for surface intersections, which can be computed algebraically. Straight lines are a good approximation as long as the curvature is reasonably small. This approach is prefered for its computational speed and simplicity. In case of strong curvature one can run straight line intersections until a certain proximity to a surface is reached and then start to target the intersection candidate. The \texttt{Navigator} resolves surface candidates when entering a volume or reaching a layer. All candidates will be intersected and filtered given their bounds. The \texttt{Navigator} will then order the candidates by their distance to the current position on the trajectory and target the closest one. Note that the order of itersections is stable as long as surfaces do not overlap. This approach is fast and simple, but comes with edge cases.

\input{parts/2_developments/06_navigator-developments/figures/edge_cases.tex}

Figure \ref{fig:dev/nev/edge-cases} illustrates a few of these edge cases. The trajectory can (a) bend away from the candidate after the straight line intersection was successful. This is handled in the \texttt{Navigator} by reintersecting the candidate after each step and discarding it if it becomes invalid. Similar to this case, the trajectory can (b) bend towards the candidate after the straight line intersection was unsuccessful. This case is not handled in the \texttt{Navigator} and will result in a missing hit. One way to improve this is to increase the surface boundary while resolving candidates. The third case (c) is a combination of the first two. While approaching a volume boundary, the trajectory can bend away from one of the boundary surfaces towards the other. This case needs to be handled correctly because the \texttt{Navigator} relies on knowing in which volume it is. This case is handled by detecting that no boundary surface is reachable anymore and resolving the new volume using the current global position.

\section{A streamlined navigation model for ACTS}

ACTS has a generic navigation implementation simply called \texttt{Navigator}. It is a longstanding component and is reasonably optimized. During the work on this thesis the \texttt{Navigator} has been reworked to improve stability, readability, and maintainability, while its computational performance stayed the same.

The \texttt{Navigator} makes use of the hierarchical structure of the tracking geometry which is composed of volumes, layers and surfaces. On initialization, the \texttt{Navigator} determines the current volume and layer given the current position on the trajectory. If no current layer is found, the \texttt{Navigator} will try to find the closest layer in the current volume. Given a layer is found, the \texttt{Navigator} will resolve the surface candidates which is accelerated by a surface grid and a spatial index given the local position of the trajectoy on the layer. The surface candidates are then intersected using a straight line from the current position and assuming the direction of the trajectory. Intersections are discarded if they are not within the boundary of the surface and sorted by their path length. The remaining candidates are targetted one after the other and hits are recorded. The \texttt{Navigator} switches to the next layer after all surface candidates have been processed. When all layers inside a volume have been processed, the \texttt{Navigator} will try to reach a volume boundary and switch to the next volume.

While this approach has proven to work well and is reasonably fast due to acceleration structures and an efficient use of intersections, there are a few caveats. This strategy is optimized for reasonably straight trajectories within a volume. It will definitely break down for looping tracks and result in missing hits for tracks with strong curvature. Given these harsh conditions experiments will have to implement a custom navigator. A more robust strategy could be to only resolve candidates given a certain proximity to a surface and otherwise step blindly forward with a bound step size. This will come with a computational performance penalty, therefore there is not a one-size-fits-all solution, if experiments depend on it.

Since a straight line is only an approximation of the trajectory, the prediction of an intersection with a surface can become invalid or be discarded wrongly in the first place. The \texttt{Navigator} in ACTS will discard intersections that become invalid and move on to the next target. While candidates which were not resolved in the first place will be lost in any case. This might be a good strategy for an experiment which depends on fast navigation and can tolerate a few missing hits, but others might favor a more robust navigation over computational performance.

During the work on this thesis the \texttt{Navigator} has been steamlined over multiple iterations. One major improvement was the correct handling of intersection ambiguities and to only discard them when enough information is available. This is covered in more detail in the following section. Furthermore, the \texttt{Navigator} has been reworked to work in concrete steps and not recover from all possible states in multiple places in the code. This makes the code more readable and maintainable and potentially slightly faster. The navigation in ACTS is now decoupled from the stepper and only receives the current position and direction of the trajectory. The surface tolerance has been synchronized and can now be set via an option on the propagation.

An issue that was not addressed during the work on this thesis, but identified as such, is the handling of a target surface during navigation. The target surface is a concept in ACTS which is used to abort the propagation if a certain surface is reached. This surface may not be part of the tracking geometry, and as such not discoverable by the navigator. Since this is an important feature for track reconstruction, a \texttt{SurfaceReached} actor exists which behaves like a second navigator in charge of trying to reach the target surface. Since these two navigators are not synchronized, they will both try to set a step size to reach their individual targets leading to propagation failures. In the future, the \texttt{SurfaceReached} actor should be removed and the navigation should be responsible for reaching the target surface.

\section{Robust intersection handling for cylinders}

The most basic surfaces ACTS deals with are planes. They are easy to intersect with a straight line and the intersection is always unique. For real detectors, plane surfaces are very common for sensitive parts, like silicon pixel detectors, but for volume boundaries they can be impractical. In case of cylindrical detectors, like at the LHC, boundary surfaces are cylinders and discs.

In general, cylinders can have none, one, or two intersections with a straight line. While for plane surfaces an intersection is unique and can be labled by the surface itself, for cylinders this results in ambiguities.

ACTS used to handle these ambiguities by rejecting one of the interersections immediately given a path length range and then picking the closest solution. One problem with this approach is that the solution of the intersection is not direction independent. Inverting the direction can result in picking the other intersection solution. The other problem is that this just hides the ambiguity and does not resolve it. Running the intersection again from a different point on the trajectory can result in silently picking the other intersection solution. During propagation this will result in failures which show up as wrong results or errors.

\input{parts/2_developments/06_navigator-developments/figures/cylinder_intersections.tex}

During the work on this thesis the intersection handling for cylinders has been improved by returning up to two solutions. The caller can then decide which solution then to use. To make the order of the solutions stable, they are ordered by the path length. This way the "entry" solution will always be the first one and the "exit" solution will always be the second one. In this general case intersections are unique given a surface and the intersection index. This is illustrated in Figure \ref{fig:dev/nev/cylinder-intersections}.

We should note one more sublety of cylinder surfaces. While the boundary tolerance will generally increase the extend of a surface in ACTS, it will not increase the radius of a cylinder. This might be unexpteced and can lead to missing intersections but is believed to be the correct solution geometrically. Currently the boundary tolerance is purely two dimensional and does not take the orthogonal direction of the surface into account. This is a limitation of the current implementation and should be addressed in the future.

\section{Removal of path limit and overstep limit}

The path limit and overstep limit express a range of path lengths the navigator is supposed to take into account. These limits have been renamed to \textit{far limit} and \textit{near limit} respectively. As described in the previous section, these limits were mainly used to resolve intersection ambiguities with cylinders. Additionally, the path limit was used to hide intersection candidates which are further away than the target surface. This helped to stabelize the navigation for inward propagations in cylindrical geometries.

After the ambiguity of cylinder intersections has been resolved inside the navigator, rather than during surface intersection, the near and far limit lost their purpose and were removed from almost all places in ACTS. The near limit is now used during surface candidate resolution to discard candidates which are too close to the current position on the trajectory. This avoid resolving candidates multiple times and cyclic navigation between two surfaces. ACTS already has a tolerance for the purpose of deciding if a global position is on a given surface. The near limit is set to this surface tolerance, which is 0.1 μm by default.

\section{Decoupling from the stepping}

Even though the navigation and the stepper are two separate concepts and components in ACTS, they used to be tightly coupled. The navigation functions would receive the stepper and its state as an input and use it to determine the position and direction of the propagation when necessary. This allows for a lot of customization, but it comes with the cost of complex constrol flow and a lot of boilerplate code. The navigation and the stepper are two separate concepts and live in different components and should be treated as such. The navigation should only be concerned with finding the next surface and the stepper should only be concerned with propagating the track parameters for a certain distance. The navigation should not have to know about the internals of the stepper and vice versa. This decoupling makes the code easier to understand and maintain and it clearly states the responsibility of the components with a concrete interface.

Since stepper and navigator do not have a virtual interface and are accessed via their concrete type in the propagation, interfaces between stepping and navigation required templating on a function level to avoid recursive template instantiation. This forced stepper and navigator code to reside in header files which ultimately drives up compile times.

The input to the navigation was trimmed down to the current position and direction of the propagation. The navigation will then return the next surface the track will intersect with. The stepper will follow the trajectory until the surface is reached and the navigation will be called again.

\section{Developments on the DirectNavigator}

Direct navigation is a technique to side step the surface candidate resolution from the tracking geometry and directly target a pre-defined sequence of surfaces. This is useful in cases where the surface candidates are known before the propagation, and CPU time and navigation stability are a concern. The \texttt{DirectNavigator} is a long-standing feature in ACTS and can be used for refitting purposes. It evolved during the work on this thesis and was improved to be more flexible and easier to use. As such, it now supports reverse navigation and automatically detects if the start surface is part of the surface sequence.

\section{Implementation of try-all navigators}

The \texttt{Navigator} has been introduced as a fast and simple navigator for reasonably straight tracks. It tries to balance computational and physics performance. As such it will not always find all the surfaces the trajectory intersects and therefore miss hits which are unrecoverable. Ultimately, an experiment needs to be able to assess the quality of the navigation. For this purpose a try-all navigator can be used. The idea is to resolve all surface candidates in a volume and intersect them continuously during the propagation, to be sure no surface was missed. As such, a try-all navigator is a reference navigator for validation purposes. It is purely optimized for accuracy and almost guarantees that no surface will be missed.

During the work on this thesis two try-all navigators have been conceptualized and implemented in ACTS. The surface candidate resolution is the same for both, they query and test all sensitive, passive and boundary surfaces in the current volume.

The \texttt{TryAllNavigator} will target the closest surface candidate and redo the candidate resolution after each step to minimize the chance of missing a surface.

The \texttt{TryAllOverstepNavigator} will order blind steps forward and resolve navigation target between start and end point of the step. This appraoch offers a better approximation of the track path and increases stability in case of strong curvature.

These two concepts can be extended to an optimized navigator and combined with regional navigation candidate resolution using acceleration structures to obtain a fast and accurate navigator for high curvature tracks.

A potential future improvement would be to combine the two try-all navigators into one and to add a toggle for the overstep resolution mode to resolve potentially missed candidates.

\input{parts/2_developments/06_navigator-developments/tables/tryall_performance.tex}

Table \ref{tab:dev/nav/tryall-performance} shows the performance of the implemented try-all navigators compared to the \texttt{Navigator} with $10^4$ single muons in the Open Data Detector using fast track simulation. The maximum step size for the try-all navigator runs was limited to 1 cm. The try-all navigators are significantly slower than the \texttt{Navigator} but are able to recover more hits. This also validates the \texttt{Navigator} and shows its efficiency in terms of physics and computation.

\section{Conceptualized solution for surface ambiguity}

During reconstruction we have to deal with uncertainties because track parameters are only known up to a certain precision. Conceptually, these uncertainties can be used during surface intersections and navigation to estimate the propability a given surface and a trajectory intersect each other. ACTS does not deal with intersection uncertainties but a surface boundary tolerance can be used to model these uncertainties. The surface boundary tolerance is a two dimensional tolerance which is added to the surface bounds. This effectively enlarges the surface and the navigator will find additional candidates that might become valid during the approach.

Using a surface boundary tolerance is just an approximation and has its limitations. As metioned before, the boundary tolerance operates in two dimensions and will for example not increase the radius of a cylinder. The other limitation is that intersection uncertainties can lead to reordering of the candidates during the approach which is not handled by the \texttt{Navigator}. This can lead to wrong results as the causal order of intersections is not respected and the propagation would have to reverse direction to reach all the surfaces.

Another problem is that coincident surfaces which are separated spacially without boundary tolerance can overlap given a tolerance. Generally, this is a geometry overlap, which results in undefined behaviour in ACTS. The path length for both surfaces will be the same and the \texttt{Navigator} will pick one of them, practically at random. To resolve this issue the propagation would have to allow hitting both surfaces at the same time. This is not supported by the current implementation but under investigation as a future improvement.
