Track finding is one of the key steps in track reconstruction. It is responsible for identifying hits that belong to the same particles. A perfect track finding algorithm can associate all hits to unique tracks corresponding to the particles that traversed the detector. In reality, however, there are effects which prevent the algorithm from finding all tracks. Particles might create too few hits due to decay, their momentum being too small, or being stopped by the material they traverse. This will result in hits that cannot be associated with a track. At the same time, detectors are not perfect and will have dead modules, resulting in missing hits, and noise that will create fake hits. These additional hits can result in tracks that were not created by a single particle, so-called fake tracks. In summary, a good track finding algorithm needs to deal with these issues and find as many tracks as possible while keeping the number of fake tracks as low as possible.

Associating hits to tracks is a combinatorial problem. The number of possible combinations grows superlinearly with the number of hits. For busy events with many hits, the track finding step in a reconstruction chain is usually the most expensive one in terms of compute time. Therefore, it has to be well optimized and balanced between physics and computational performance.

ACTS provides a Combinatorial Kalman Filter (CKF) implementation which can be used as a component for the track finding step in an experiment. The CKF depends on an initial estimate of the track parameters, which can be obtained from a prior seeding step. It searches iteratively for compatible measurements and updates the track parameters and their covariance using the Kalman filter formalism. Given measurement ambiguity, the CKF can branch into multiple hypotheses.

While ACTS already had a working prototype implementation for the CKF, it has been effectively rewritten during the work on this thesis to improve its performance, flexibility, and maintainability. The new implementation and the most important changes are presented in this chapter. First, the CKF algorithm and its implementation is described in detail. Then, the revision of the CKF into a track finding component is discussed. The branch stopper extension is presented which allows to stop the track search early. Modeling of branches with the track EDM is discussed and the branch finalization is explained. Two techniques to improve the track finding performance are presented: seed deduplication and stay-on-seed strategy. A two-way track finding strategy is introduced which allows to recover all measurements independent from the seed position. Finally, the chapter is concluded with a discussion of the initial covariance and the shortcomings of the CKF implementation.

\section{The \texttt{CombinatorialKalmanFilter}}

\input{parts/2_developments/07_trackfinding-developments/figures/ckf_flow.tex}

The \texttt{CombinatorialKalmanFilter} in ACTS is a Core algorithm that uses initial track parameters to iteratively find compatible measurements and fit them on a track.  As such, it relies heavily on the propagation of track parameters and their covariance through the tracking geometry and magnetic field to find all the sensitive surfaces the track traverses. When the CKF arrives on a surface, it calls the \texttt{TrackStateCreator} to retrieve all compatible measurements given the predicted track parameters and covariance on the surface. In case of multiple compatible measurements, the CKF creates branches for each of them. These branches are processed in a depth-first search manner. The \texttt{TrackStateCreator} is provided by the experiment as an interface for ACTS to access the measurements. This step might also perform a calibration of the measurement given the track hypotheses. The newly created track states are then used to update the track parameters and their covariance using the Kalman filter formalism. In case no compatible measurements are found, the hole count of the track is increased. If material is encountered, the CKF will compute the mean energy loss and multiple scattering to update the track parameters and their covariance. The material update is split into a pre- and a post-update which allows users to divide the material amount into parts before and after reaching the surface. As a final step, the \texttt{BranchStopper} is called to decide whether the CKF should continue or stop the search. This procedure is repeated until the CKF reaches the end of the tracking geometry, the path limit is reached, or the branch stopper decides to stop. The resulting tracks are then stored and returned to the user. Figure \ref{fig:dev/trackfinding/ckf-flow} illustrates the described flow of the algorithm.

The \texttt{TrackStateCreator} is a new abstraction which bundles source link access, measurement selection and calibration together. An implementation for a track state creator in ACTS Core exists, which calls these individual components if the user does not provide a custom implementation. The \texttt{SourceLinkAccessor} is responsible for accessing the measurements on a surface. The \texttt{MeasurementSelector} is responsible for selecting the measurements which are compatible with the predicted track parameters. The \texttt{Calibrator} is responsible for calibrating the measurements given predicted track parameters. The reason for this new abstraction is to allow for optimizations across all three components. It was observed that the track state creation had a significant impact on the CKF performance and that the components could not be optimized individually.

Since the Kalman filter is optimal for linear systems, the CKF has an optimal discrimination power for selecting measurements. Measurements are usually selected based on their compatibility with the predicted track parameters using a $\chi^2$ test. Given the degrees of freedom, the user effectively selects a p-value threshold to decide whether a measurement is compatible with the track or not.

The CKF will provide optimal resolution on the last track state that has a measurement. A smoothing pass can be performed, to inform the other track states about all the measurements. This results in fully fitted tracks and optimal track parameter estimates. Users may want to refit tracks given an estimated particle hypothesis. Especially in case of electrons as the Kalman filter will not model the energy loss correctly.

The usage of the \texttt{CombinatorialKalmanFilter} is exercised in the Examples with the \texttt{TrackFindingAlgorithm}. Given a configuration and estimated track parameters, it will try to find all tracks and store them as intermediate event data. Another algorithm can then refit these tracks or perform a vertex search.

\section{CKF as a component}

The \texttt{CombinatorialKalmanFilter} used to be a monolithic algorithm with multiple responsibilities. It used to deal with finding compatible measurements, fitting the track parameters, smoothing the track, and extrapolating the track to a reference surface. The code was hard to reason about and the error handling was not done properly due to the high complexity. Because all these different tasks were coupled, a problem during the smoothing or extrapolation step would silently affect the finding efficiency.

An important missing piece to the CKF was the ability to find measurements in both directions. The CKF was only able to find measurements in the particles' momentum direction. Since seeds are not required to be formed on the innermost layer of the detector, measurements on such inner layers might be missed which results in degraded efficiency and resolution. Adding this feature to the CKF would have further complicated the code and made it even harder to debug and maintain.

To address these issues, the \texttt{CombinatorialKalmanFilter} has been reduced to a pure track finding component. Smoothing and extrapolation have been extracted to toolbox functions. A track finding algorithm can call these if necessary and at the right time. This way, the experiment is in control of the track finding process and can directly steer when to spend precious compute time.

\section{Branch stopper}

Seeding algorithms can produce many seeds which will not result in a track. The CKF will then spend compute time to propagate through the tracking geometry, only to find out that the track has too many holes and too few measurements.

A \texttt{BranchStopper} can be used to stop the CKF if it is clear that no track will be found. The \texttt{BranchStopper} is a user provided function which can observe the current track and its track states to make a decision. There are three possible outcomes: \textit{continue}, \textit{stop and drop}, and \textit{stop and keep}. Where \textit{continue} will continue the CKF as usual; \textit{stop and drop} will stop the CKF and drop the branch; and \textit{stop and keep} will stop the CKF and keep the branch.

\section{Track container for branches}

Branches discovered during running the CKF are now stored as tracks in a track container. The reason for this is that CKF branches effectively resemble tracks. They have a measurement count, a hole count, an outlier count, and a $\chi^2$ sum. The track container is generic and allows experiments to add additional information through dynamic columns. An immediate benefit of this approach is that the branch stopper can now accumulate experiment specific statistics about the track, for instance the number of pixel subsystem holes, and make a more informed decision about whether to keep or drop a branch.

\section{Branch finalization}

The branch finalization is triggered when a propagation abort condition is met or a branch is stopped by the user. The current branch is then stored in the output track container, given it is selected to be kept. If other branches are still active, the CKF will continue with the next one, otherwise the CKF will finish.

Finalizing a branch was previously done in multiple places and was essentially duplicated. It has been refactored to a single function which can be called from different places. This results in better maintainable code which is easier to reason about.

\section{Seed deduplication}
\label{sec:dev/trackfinding/seed-deduplication}

Seed deduplication is a technique to avoid finding the same tracks through different seeds multiple times. For instance, a triplet seeding algorithm can produce many seeds for the same particle. While this is partially by design to gain redundancy and to mitigate operation effects, e.g. inactive modules, it can also be a waste of compute time.

The seed deduplication technique has been implemented in the \texttt{TrackFindingAlgorithm} in the ACTS Examples and can be toggled in the configuration. After a track has been found, its measurements are checked against the measurements of all seeds. Seeds which are fully incorporated in the found track are flagged and will be skipped in the following track finding iterations. This implementation has the benefit of only checking the deduplication when a track is found. There is no additional computational cost for dropped fake tracks. The current implementation will iterate over all possible combinations of measurement triplets on a track, and flag corresponding seeds using a hash table. While the hash table seems to be an efficient solution, the loop over the combinations might be suboptimal and can be improved in the future.

\input{parts/2_developments/07_trackfinding-developments/tables/seed_deduplication.tex}

Table \ref{tab:dev/trackfinding/seed-deduplication} shows the performance of the track finding algorithm with and without seed deduplication. The results were obtained with the OpenDataDetector, \ttbar events at $\sqrt{s} = 14 \ \text{TeV}$, $\langle \mu \rangle = 200$, using fast simulation and a standard reconstruction workflow. It shows that the deduplication has a significant impact on the time per event, but only a small impact on the track efficiency.

\section{Stay-on-seed strategy}
\label{sec:dev/trackfinding/stay-on-seed}

Stay-on-seed is a strategy to keep the CKF on the selected measurements of the seed and prohibit it from branching into other measurements. Since the initial parameter covariance is inflated, the CKF will see a couple of compatible measurements in the vicinity of the seed. In case of a triplet seeder, the combinatorics of these measurements are already explored and the CKF does not need to redo this work.

This strategy has been implemented in the \texttt{TrackFindingAlgorithm} in the ACTS Examples and can be toggled in the configuration. When this feature is turned on, the source link accessor is decorated to only return the seed measurements if the CKF is on one of their surfaces.

\input{parts/2_developments/07_trackfinding-developments/tables/stay_on_seed.tex}

Table \ref{tab:dev/trackfinding/stay-on-seed} shows the performance of the track finding algorithm with and without stay-on-seed. The results were obtained with the OpenDataDetector, \ttbar events at $\sqrt{s} = 14 \ \text{TeV}$, $\langle \mu \rangle = 200$, using fast simulation and a standard reconstruction workflow. It shows that the stay-on-seed strategy has a significant impact on the time per event, but only a small impact on the track efficiency.

We should note that this will only make a difference if branching is enabled and the hit density is high enough to result in multiple compatible measurements in the vicinity of the seed. In the example shown, the hit density is high and branching was set to a maximum of 10 per surface.

\section{Two-way track finding}
\label{sec:dev/trackfinding/two-way}

In order to find all measurements a particle has left in the detector, the CKF either needs to start at the very first measurement or needs to be able to look for measurements in both directions. Once in the forward direction, the particle traveled along, and once in the opposite direction, towards the origin of the particle. Since the CKF only branches in the propagation direction, this cannot be simply added to the existing algrithm.

After decoupling the CKF from the smoothing and extrapolation, it can be used to run twice: once with the initial track parameters from the seed, and a second time in reversed direction with the smoothed parameters at the first measurement. This has first been implemented and validated with the \texttt{TrackFindingAlgorithm} in the ACTS Examples, and later in the ATLAS reconstruction software.

The remaining challenge is to stitch the resulting track segments together. The ACTS track EDM does not dictate a preferred ordering for the stored tracks. But it does also not provide a way to express the ordering of a track. For this reason, tracks should always be stored with a consistent ordering. For this reason, track segments pointing in the opposite direction need to be reversed before they are concatenated to a full track. This involves an inverse of the bound-to-bound Jacobian which is stored on each track state and used for the smoothing.

\input{parts/2_developments/07_trackfinding-developments/tables/two_way.tex}

Table \ref{tab:dev/trackfinding/two-way} shows the performance of the track finding algorithm with and without two-way search. The results were obtained with the OpenDataDetector, \ttbar events at $\sqrt{s} = 14 \ \text{TeV}$, $\langle \mu \rangle = 200$, using fast simulation and a standard reconstruction workflow. It shows that the two-way search is able to reduce the duplication rate by deduplicating seeds more effectively. Even though the two-way search involves more computation, it is slightly faster as fewer seeds have to be processed.

\section{Initial covariance}

The Kalman filter depends not only on predicted parameters, but also on their covariance to form a filtered state given a measurement. The same is true for the initial parameters and their covariance. To perform an unbiased fit to the measurements for both, the parameters and their covariance, the initial covariance needs to be inflated to express little prior knowledge about the track \cite{fruehwirth}. In the case of the CKF, the covariance will not only be used to filter the measurements, but also to select them. Given an inflated covariance, the CKF will be able to pick practically any measurement on a sensor since the $\chi^2$ will be effectively scaled down by the inflated covariance.

For this reason, the initial covariance needs to be scaled properly by the experiment. If branching is enabled and the hit density is high, an inflated covariance will result in a large search range and the CKF will spend a lot of compute time following all possible branches. If the covariance is too small, the CKF might miss measurements which would have been compatible with the track. At the same time, the estimated parameters and their covariance will be biased.

Given the ACTS bound track parameterization, described in Section \ref{sec:intro/acts/track-parameterization}, $\mathbf{x}_\text{local} = \left(l_0, l_1, \phi, \theta, q/p, t\right)^\mathrm{T}$, it was found that the initial covariance can be set to a diagonal matrix with constant values for all parameters except $\lambda = q/p$. The local position (and if given, time) covariance will be effectively set by the resolution of the first module, the CKF starts its search on. Similarly, the angle covariance will be dominated by two consecutive spatial measurements. The $\lambda$ variance cannot be put to a constant, because the momentum resolution of a detector is usually relative to the momentum itself, $\sigma_{p_T}/p_T \approx \text{const}$, and the conversion from $p_T$ to $p$ introduces a $\theta$ dependance. Explicitly, the $\lambda$ variance can be dimensioned with
\begin{equation}
    \sigma^2_\lambda = \left( \lambda \, \frac{\sigma_{p_T}}{p_T} \right)^2 + \left( \frac{\lambda}{\text{tan}\,\theta} \, \sigma_\theta \right)^2 \text{.}
\end{equation}

\section{Shortcomings}

While a lot of progress has been made on the CKF during the work on this thesis, and it shows promising physics and computational performance, there is still room for improvements.

\begin{description}
    \item[Generic track state creator] The \texttt{TrackStateCreator}, which is provided by the user to the CKF as an interface to access the measurements, is a component with common functionality between experiments. ACTS could provide a generic implementation with customization points for this purpose. This would reduce implementation effort for the experiments and allow sharing improvements between them.
    \item[Extended bounds] The CKF is currently only able to find measurements on surfaces the predicted trajectory strictly intersects. This is a limitation if a measurement lies close to the boundary of the surface and the predicted trajectory misses the surface with a small distance. To mitigate this, the surface bounds could be extended during navigation to allow for a search range around the predicted trajectory. This requires small modifications of the \texttt{Navigator} and proper handling in the CKF to avoid counting holes if sensor was hit on the extended bounds but no measurement was found.
    \item[Edge holes] A similar effect that needs to be addressed are edge holes. If the predicted trajectoy barely intersects a sensor but no measurement is found, the CKF will incrase the hole count even though the actual particle trajectory might not have hit the sensor. This can be mitigated by defining a region around the sensor boundary where holes are not counted. Work has been done to explore this idea inside the ITk tracking group but has not been fully implemented in ACTS yet.
    \item[Multiple compatible surfaces] A limitation of the CKF implementation is that it will only try to find measurements on a single surface at a time. For instance, this will have implications for the ITk reconstruction because some sensitive surfaces have boundaries which "touch". Given inaccuracies during reconstruction a neighboring surface might be picked and the actual measurement of a particle will not be recovered. This can be mitigated by allowing the CKF to search for measurements on multiple surfaces at the same time, which also requires modifications to the \texttt{Navigator}.
    \item[Multiple compatible surfaces \#2] The former problem could also be viewed in a broader context. The CKF is currently bound to intersect a single surface at a time. Potential branching can only happen on the surface in presence of multiple compatible measurements. This is a limitation if the predicted trajectory has a high uncertainty and multiple surfaces are compatible with the predicted trajectory. In this case the CKF should branch before choosing one of the surfaces. How the \texttt{Navigator} should handle this and how much communication would be necessary between the CKF and the \texttt{Navigator} is unclear. It should be noted that this limitation has a negligible impact on typical high energy physics experiments since tracks are usually stiff and the uncertainty small compared to the sensor dimensions. 
\end{description}
