The High-Luminosity (HL) upgrade \cite{hl-lhc-tdr} of the Large Hadron Collider (LHC) at CERN will substantially increase its instantaneous luminosity, leading to a greater number of collisions per bunch crossing. In this environment, computational performance of reconstruction algorithms is crucial and Acts has been chosen to meet the challenges posed by high pileup conditions.

This thesis has shown the latest developments and performance of the Acts toolkit \cite{acts-2022} --- a community-driven project that provides experiment-independent tracking algorithms written in modern C++. The developments include improvements to the general infrastructure including the numerical intergration and detector navigation (Chapter \ref{ch:dev/general}, \ref{ch:dev/stepper}, and \ref{ch:dev/navigator} respectively), track finding (Chapter \ref{ch:dev/track_finding}), propagation in dense material (Chapter \ref{ch:dev/dense}), and 4D tracking (Chapter \ref{ch:dev/4d_tracking}). The performance of Acts was evaluated using the OpenDataDetector (ODD) \cite{odd-v2} (Chapter \ref{ch:perf/odd}) and the ATLAS \cite{atlas} Inner Tracker (ITk) \cite{itk-pixel-tdr}\cite{itk-strip-tdr} (Chapter \ref{ch:perf/itk}) in terms of computation and physics.

One of the main achievements was the development of a general purpose track finding algorithm and application to the ITk, which will be used for reconstruction after the Phase-2 upgrade during Run 4 of the LHC. The track finding algorithm has been carefully optimized to ensure it can handle the increased event complexity expected at the HL-LHC, within a tight CPU budget while maintaining high efficiency and low fake rates.

The developments on 4D tracking provide a promising outlook for the future of tracking in high-energy physics experiments with new emerging fast timing detector technologies. The integration of time information into reconstruction has the potential to significantly improve physics performance by reducing the impact of pileup. Chapter \ref{ch:dev/4d_tracking} summarizes the latest developments in Acts for 4D tracking. The performance was shown as part of Chapter \ref{ch:perf/odd} with the ODD.

All these developments and performance evaluations are not bound to a single experiment, but can be applied to any high-energy physics experiment. The modularity and flexibility of Acts allows for easy adaptation to different experimental setups and requirements. This makes Acts a valuable tool for the future of high-energy physics.
