This chapter will discuss general developments on ACTS, performed during the course of this thesis. They are not specific to any experiment or use case but are general improvements to the software. First, developments on the propagation component are discussed. Then, a new measurement projection mechanism is presented, which provides a fast and generic way to compare predictions from the propagation to measurements. Afterwards, developments on the examples framework are discussed, followed by the implementation of a greedy ambiguity resolution algorithm, the implementation of a particle hypothesis, the redesign of the boundary tolerance, and the implementation of the modified Bryson-Frazier (mBF) smoother.

\section{Propagator developments}

The \texttt{Propagator} is a central component in ACTS and is used by various reconstruction algorithms. It is responsible for transporting track parameters through the geometry. This task is subdivided into stepping and navigation. The stepping is responsible for moving the track parameters along the trajectory of the particle and the navigation is responsible for finding and targeting the next geometry object, the trajectory intersects. The propagation is responsible for the overall control flow and handles the communication between stepping and navigation. Stepping and navigation developments are discussed in more detail in Chapter \ref{ch:dev/stepper} and \ref{ch:dev/navigator}, respectively.

During the work on this thesis several improvements were made to the \texttt{Propagator}:

\begin{description}
    \item[Propagation options] The \texttt{PropagatorOptions} were reworked to provide a more consistent and flexible interface. They have been subdivided into \texttt{StepperOptions} and \texttt{NavigatorOptions} to separate concerns. Apart from that, the base options now contain all the necessary information to configure the propagation for different use cases, for instance dense material propagation.
    \item[State creation and initialization] The creation and initialization of the \texttt{PropagatorState} was reworked and elevated to the public interface. This allows usage of a single state object for multiple propagation calls which can be beneficial for specific algorithm workflows and their performance.
    \item[Reworked propagation loop] The propagation loop was reworked after isolating the concerns of stepping and navigation. The \texttt{Propagator} now handles the control flow and communication between the stepping and navigation explicitly.
    \item[Merging of actor and aborter] Actors and aborters allow to steer the propagation by injecting custom behavior into the propagation loop. These two concepts were merged into one to simplify the interface and to allow direct communication between them.
    \item[Alternative steering mechanism] An alternative steering mechanism was conceptualized which allows algorithms to steer the propagation without actors and aborters. It is discussed in more detail in the following Subsection \ref{subsec:dev/general/alternative-steering}.
\end{description}

\subsection{An alternative steering mechanism}
\label{subsec:dev/general/alternative-steering}

Reconstruction algorithms which depend on the propagation are currently implemented as \texttt{Actor}s. They are part of the propagation loop and called after each step. While this mechanism is sufficient to host all current algorithms, it has the disadvantage of tight coupling to the propagation. With this design, the propagation ends up steering the control flow of reconstruction algorithms. Instead of the algorithm asking the propagation to reach for the next sensitive surface, the propagation is instructed to traverse the geometry and call the algorithm's \texttt{Actor} after each step.

Under these circumstances, trying to understand what an algorithm is doing also requires to understand what the propagation is doing. During the work on this thesis numerous improvements and fixes were made to the propagation, the stepping, and the navigation. This required a deep understanding of these concepts and their implementation. Moreover, it was immediately clear that improving the performance of the track finding, which is the main focus of this thesis, could only be possible by also improving the propagation.

As a result of this experience a new steering mechanism for the propagation was conceptualized. The idea is to invert the control flow and let the algorithm steer the propagation. This would decouple the algorithm from the propagation loop and increase readability and maintainability. The algorithm would instruct the propagation when to stop beforehand and when to continue. The propagation would only be responsible for the geometry traversal and the algorithm would be responsible for the track finding for instance.

The \texttt{Propagator} was modified to allow for the creation of a \texttt{PropagatorState} object by the user which can then be used to start and continue the propagation while an \texttt{Aborter} can be used to stop the propagation at any time. An algorithm can use this mechanism to propagate from one surface to another and deal with the reached surfaces outside the propagation loop. A next step would be to implement an algorithm like the Kalman track fitter using this mechanism and compare the performance and readability to the current implementation. This was not done as part of this thesis but is an interesting direction for future work.

\section{A new measurement projection mechanism}

Measurement projection is a mechanism in ACTS to allow comparing predicted track parameters to measurements in a generic way. In ACTS, measurements are represented as a subset of bound track parameters, which are defined as $\mathbf{x}_\text{local} = \left(l_0, l_1, \phi, \theta, q/p, t\right)^T$. This is a simple and powerful abstraction which allows to represent any combination of measurement dimensions.

This projection used to be a matrix multiplication of the track parameters with a projection matrix. The projection matrix is provided by the measurement, along with the measurement parameters and covariance. During optimizations on the track finding component it became clear that this implementation has overhead due to the matrix multiplication.

During the work on this thesis a new measurement projection mechanism was implemented. Instead of using a matrix multiplication, the measurement simply carries an array consisting of the bound parameter indices it contains. This was consitently adapted throughout the code base including the track EDM, the Kalman filter, and the Examples framework. The new mechanism is faster and more memory efficient than the old mechanism.

\section{Developments on the Examples framework}

The Examples framework is used to demonstrate the capabilities of ACTS, to validate the correct implementation of the Core algorithms, and to provide a starting point for new experiments. It consists of IO modules, EDM implementations, and example algorithms which can be composed to build a full simulation and reconstruction chain.

To validate and improve the track finding in ACTS Core, the Examples framework was revised and improved over multiple iterations during the work on this thesis. This work is beneficial to other experiments while they prototype the detector design or the reconstruction chain. The following developments were made to the Examples framework:

\begin{description}
    \item[Aliasing mechanism] An aliasing mechanism for the \texttt{Whiteboard} and \texttt{Sequencer} was implemented which allows to modularize reconstruction chains. For instance, tracks may need ambiguity resolution or refitting after they are found. The vertexing depends on a track collection as input but will not know about the ambiguity resolution or refitting. The aliasing mechanism allows routing for these different track collections by effectively stating which is the most "recent" one.
    \item[Event generation] Event generation in ACTS Examples offers a particle gun and Pythia8 \cite{pythia8} . Both have been improved in terms of parameterization and to export truth information about the generated vertices.
    \item[Geant4 integration] The Geant4 integration for simulation was improved in terms of stability and maintainability by reworking the life cycle. The Geant4 integration allows to simulate the detector response and the particle interactions in the detector.
    \item[Measurement container] The measurement has been rewritten to improve memory density and performance. It used to consume a full bound parameter dimension for each measurement. The new implementation stores jagged arrays without any overhead.
    \item[Central track truth matching] The track truth matching used to be done in various places where it did not necessarily behave the same. A central track truth matching was implemented to ensure consistency and to provide a single point of access for the truth information.
    \item[Vertexing developments] Studying the vertexing performance required to effectively rewrite the \texttt{VertexNTupleWriter}, which is responsible for writing the reconstructed and truth matched vertices to an output file. The output now also includes metrics like vertex density, track contamination, split rates, and merge rates. Additionally, the \texttt{TruthVertexSeeder} was implemented to allow measuring the finding performance independent from the seeder performance.
\end{description}

\section{Implementation of \texttt{GreedyAmbiguityResolution}}
\label{sec:dev/general/greedy-ambiguity-resolution}

Track finding returns a track candidate container that can contain duplicate and fake tracks. These ambiguities have to be resolved to get a clean track collection which represents the event as good as possible. This is the responsibility of the ambiguity resolution and it is a crucial step in the tracking workflow which can be done in various ways. The greedy ambiguity resolution is a simple and fast method to resolve ambiguities. It is based on the idea that the best track candidate is the one with the most hits and least shared hits. If two competing track candidates share hits, the one with the best $\chi^2$ is chosen. This process is repeated until no ambiguities are left.

This procedure was implemented in ACTS as \texttt{GreedyAmbiguityResolution} during the work on this thesis. The motivation for this implementation was to study vertexing performance with the OpenDataDetector. Vertexing depends on a clean track collection and the greedy ambiguity resolution was the easiest way to achieve this. The implementation is generic and can be used with any track collection.

The user can provide a custom track selector to define the quality cuts and a custom track comparator to define the order of tracks. The track selector is used to filter out tracks that are not considered for the ambiguity resolution. The track comparator is used to compare two tracks and decide which one is better. Users can provide a custom track selector and track comparator to specialize the behaviour of the ambiguity resolution.

\section{Implementation of \texttt{ParticleHypothesis}}

The particle hypothesis is an important concept in reconstruction. Usually we do not know the particle type which generated a track. But in order to reconstruct the track, we have to make an assumption to predict the trajectory of this unknown particle. These assumptions can be summarized as a \texttt{ParticleHypothesis} and provided by the user. The information is stored along track parameters and inside track containers to record these assumptions.

Before this implementation, ACTS was missing a central definition of a particle hypothesis. Parameters like mass and charge were inputs to different modules and algorithms, but there was no consistent way to set it in a single place by the user. During the work on this thesis, \texttt{ParticleHypothesis} was conceptualized and implemented in ACTS.

The goal of this implementation was to be lightweight and flexible to align with the experiment independent design of ACTS. The material interaction model of ACTS only requires the mass, charge, and particle type. To avoid maintaining a full list of particles it was preferred to keep these parameters generic and only have a few common particles predefined. The user can define their own particle hypotheses by providing the mass, charge, and particle type. The particle type is modeled as an integer holding the absolute PDG code of the particle. The sign of the PDG code encodes the sign of the charge. Since the material interactions in ACTS are symmetric with respect to the charge sign, this information can be omitted. The mass and charge are stored as \texttt{float}s.

\section{Redesign of \texttt{BoundaryTolerance}}

\texttt{BoundaryTolerance} is a powerful concept in ACTS which is not utilized to its full extend yet. Each \texttt{Surface} holds \texttt{SurfaceBounds} which define the boundary of the surface. During propagation, the trajectory of a particle is approximated by a straight line. This approximation is only correct if no forces act on the particle. In reality, the trajectory will be bent by the magnetic field or redirected by scattering. To account for these effects during propagation, one can enlargen the surface boundary using a \texttt{BoundaryTolerance}.

The concept and implementation of the boundary tolerance has been reworked during the work on this thesis. There are two parts of the implementation: the parameterization of the boundary tolerance and the implementation of the boundary check using the tolerance.

The parameterization of the boundary tolerance in ACTS can take different forms: none, infinite, absolute euclidean, and $\chi^2$ bound.

\begin{description}
    \item[None] Disables the tolerance and the surface boundary is used as it is.
    \item[Infinite] Effectively disables the boundary check.
    \item[Absolute euclidean] The tolerance is a scalar value in native length units (mm). Given a point to check, the closest point on the boundary is computed and the euclidean distance is checked against the tolerance. Only if the distance is within the tolerance, the boundary check is considered as passed. This relies on the Jacobian of the coordinate transformation and is only appropriate for small distances.
    \item[$\chi^2$ bound] Consists of a $\chi^2$ cut value and a weight matrix in local coordinates. Given a point to check, the closest point on the boundary is computed and the difference in bound coordinates is used to calculate a $\chi^2$ value using the weight matrix. Only if the $\chi^2$ value is below the cut value, the boundary check is considered as passed.
\end{description}

A boundary check with tolerance has to be provided by an implementation of \texttt{SurfaceBounds} and is used by the \texttt{Surface} for intersection checks and global to local coordinate conversions. Since the boundary check using a tolerance can be tedious to implement, two general purpose default implementations are provided: a rectangle bounds check and a convex polygon bounds check. Almost all surface bounds can be represented with these two cases in local coordinates.

\section{Implementation of \texttt{MbfSmoother}}

The modified Bryson-Frazier (mBF) smoother \cite{mbf-smoother} is an alternative to the Rauch-Tung-Striebel (RTS) smoother \cite{rts-smoother} with potential advantages in terms of numerical stability and performance. The mBF smoother does not rely on the inversion of the covariance matrix in parameter space and can reuse information from the forward filter to improve the computational performance further.

The RTS smoother is a long standing feature in ACTS and is called \texttt{GainMatrixSmoother}. The mBF smoother was implemented during the work on this thesis. It is called \texttt{MbfSmoother} and is a drop-in replacement for the RTS smoother. The user can choose between the two smoothers in the relevant tools or algorithms, for instance the Kalman track fitter.

The implementation of the mBF smoother does not reuse information from the forward filter yet. This requires additional information on the track state and error handling and is planned for future work.

Even without the reuse of information the mBF smoother was measured to be about 2 times faster than the RTS smoother in a standard track finding and fitting workflow with the OpenDataDetector. The performance gain can be attributed to the simpler matrix inversion and the accelerated handling of material states. While the RTS smoother has to perform a covariance matrix inversion for each material state, the mBF smoother does not perform any matrix inversion in this case.
